---
title: "Performance of ALD biomarkers as a function of prevalence rate"
output: html_document
date: "2025-12-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Define some functions for biomarker evaluation


## Estimate metrics from truth table

```{r calculate_metrics}

calculate_metrics <- function(TP, TN, FP, FN, metrics = c("FPR", "FNR", "PPV", "NPV", "PSI", "NNP", "NND")) {

    df = data.frame(matrix(ncol = length(metrics) + 1, nrow = 0))
    
    # Calculate PPV and NPV
    PPV = TP / (TP + FP)  # Positive Predictive Value
    NPV = TN / (TN + FN)  # Negative Predictive Value
    
    # Calculate Sensitivity and Specificity
    TPR = TP / (TP + FN)  # True Positive Rate
    TNR = TN / (TN + FP)  # True Negative Rate
    FPR = 1 - TNR
    FNR = 1 - TPR
    
    # Derive NNX
    PSI = PPV + NPV - 1
    NNP = 1/PSI
    J   = TPR + TNR - 1
    NND = 1/J
    
    # Print the metrics
    #metrics = c("P", "TP", "FN", "N", "TN", "FP")
    #cat(paste0("PREV: ", sprintf("%.4f", prevalence), "; ", paste(metrics, sprintf("%.4f", sapply(metrics, function(x) get(x))), sep = ": ", collapse = "; "), "\n"))
    df = rbind(df, as.vector(sapply(metrics, function(x) get(x))))

    names(df) = c(metrics)
    #print(df)
    return(df)
}
```

## Given TPR and TNR, create truth table and call $calculate_metrics$ above for given population size and prevalence

```{r estimate_performance}
estimate_performance <- function(TPR, TNR, prevalence, n_pop = 1e6, metrics = c("FPR", "FNR", "PPV", "NPV", "PSI", "NNP", "NND")) {

    df = data.frame(matrix(ncol = length(metrics), nrow = 0))
    
    P = round(n_pop*prevalence)  # Actual Positives
    N = n_pop - P  # Actual Negatives
    
    # Confusion matrix values
    TP = round(TPR*P)  # True Positives
    TN = round(TNR*N)  # True Negatives
    FN = P - TP  # False Positives
    FP = N - TN  # False Negatives
    
    df = calculate_metrics(TP, TN, FP, FN, metrics=metrics)
    df$PREVALENCE = prevalence
    #print(df)
    return(df)
}
```


## Given TPR and TNR, calculate metrics for a range of prevalences

```{r simulate_estimate}
simulate_estimate <- function(panel, TPR, TNR, n_pop = 1e6, maxp = 0.5, metrics = c("FPR", "FNR", "PPV", "NPV", "PSI", "NNP", "NND")) {

    # Number of estimated NNPs for regression
    N = 160
    
    # Arithmetic series
    
    Astart = 0.10
    Aend = maxp
    
    prev_a = seq(from = Astart, to = Aend, length.out = N)
    
    # Geometric series
    
    Gstart = 1e-4
    Gend = 0.1
    # 1e-3 to 0.5, in 20 steps
    # multiplier = (0.5/1e-3)^(1/20) = 1.364421
    Gmult = (Gend/Gstart)^(1/N)
    prev_g = Gstart*Gmult^(0:(N-1))
    
    # Merge series or pick one
    spectrum = c(prev_g, prev_a)
    #spectrum = prev_g
    
    df = data.frame(matrix(ncol = length(metrics) + 1, nrow = 0))
    
    #spectrum = c(1e-3, 1e-2, 1e-1, 5e-1)
    df = NULL
    for (prevalence in spectrum) {
      if (is.null(df)) {
        df = estimate_performance(TPR, TNR, prevalence, n_pop, metrics=metrics)
      } else {
        df = rbind(df, estimate_performance(TPR, TNR, prevalence, n_pop, metrics=metrics))
      }
    }
    df$Panel = panel
    #print(df)
    return(df)
}

```

## Helper functions for plotting the trends

```{r plotting-interpolation}

# Interpolate the NNP value at specific values
interpolate_values_xy <- function(df, panel, prevalence, metric="NNP", MAX_P=0.5) {
  df2 = df %>% filter(Panel == panel)
  #print(df)
  #print(paste0("filtering for ", panel))
  #print(df2)
  a = approx(df2[["PREVALENCE"]], df2[[metric]], xout = prevalence)
  ret = c(as.numeric(a$x), as.numeric(a$y))
  #print(ret)
  return(ret)
}

interpolate_values <- function(df, panel, prevalence, metric="NNP", MAX_P=0.5) {
  df2 = df %>% filter(Panel == panel)
  a = approx(df2[["PREVALENCE"]], df2[[metric]], xout = prevalence)
  return(as.numeric(a$y))
}

```

## Actual function plotting the trends

```{r plotting-trends}
# Make a list of ggplot objects
plot_metric_trends <- function(performances, prevalences=NULL, primary_prevalence=NULL, secondary_prevalence=NULL, metric="NNP", Y_MAX=20, MAX_P=0.5, draw_cv_ci=0) {

    tasks = unique(performances$phenotype)
    
    # Plot the trends
    plots = list()
    for (task in tasks) {
      #cat(task)
        my_maxp = MAX_P
        study_prevalence = NA
        if (!is.null(prevalences)) {
          x = prevalences %>%
            filter(phenotype == task & set == 'test')
          study_prevalence = x$Prevalence
          my_maxp = max(MAX_P, ceiling(study_prevalence*10)/10)
        }
        
        ####################
        # Test performances
        ####################
        
        test_params = performances %>%
                        filter(phenotype == task & set == 'test') %>%
                        select(Panel, TPR, TNR)
        df = do.call(rbind, pmap(list(test_params$Panel, test_params$TPR, as.numeric(test_params$TNR)), function(x, y, z) simulate_estimate(x, y, z, maxp=my_maxp)))

        y_max = ceiling(max(df[[metric]]))
        if (y_max > Y_MAX) {
            y_max = Y_MAX
        }

        # Add singe or two stage
        df$PanelType = ifelse(grepl("==> TE", df$Panel), "With TE followup", "Standalone test")

        # Create the plot

        # Legend for linetype only if there are multiple test types
        if (length(unique(df$PanelType)) > 1) {
            plot = ggplot(df, aes(x = PREVALENCE,
                                  y = get(metric),
                                  color = gsub(" ==> TE", "", Panel),
                                  linetype = PanelType)
                          ) +
                        labs(title = task,
                             x = "Prevalence",
                             y = metric
                          ) +
                        scale_linetype_manual(values = c("With TE followup" = "dashed",
                                                         "Standalone test" = "solid"))
        } else {
            plot = ggplot(df, aes(x = PREVALENCE,
                                  y = get(metric),
                                  color = gsub(" ==> TE", "", Panel))
                          ) +
                        labs(title = task,
                             x = "Prevalence",
                             y = metric
                             )
        }
        
        # Add vertical lines for prevalences
        if (!is.null(prevalences)) {
            plot = plot + geom_vline(xintercept = study_prevalence,
                                    linetype = "dashed",
                                    color = "#08519c"
                                    ) +
                          annotate("text",
                                   x = study_prevalence,
                                   y = min(Y_MAX, max(df[[metric]])),
                                   label = paste0("This study (", 100*round(study_prevalence, 3), "%)"),
                                   color = "#08519c",
                                   angle = 90,
                                   vjust = -0.5, hjust = 1.0
                           )
            interpolation_target = study_prevalence
        }
        # secondary is plotted before primary, because we want primary to have preference for showing interpolated values
        if (!is.null(secondary_prevalence) && !is.na(secondary_prevalence[[task]])) {
            plot = plot + geom_vline(xintercept = secondary_prevalence[[task]],
                                     linetype = "dashed",
                                     color = "#3182bd"
                                     ) + 
                          annotate("text",
                                   x = secondary_prevalence[[task]],
                                   y = min(Y_MAX, max(df[[metric]])),
                                   label = paste0("Secondary care (", 100*round(secondary_prevalence[[task]], 3), "%)"),
                                   color = "#3182bd",
                                   angle = 90,
                                   vjust = -0.5, hjust = 1.0
                                   )
            interpolation_target = primary_prevalence[[task]]
        }
        if (!is.null(primary_prevalence) && !is.na(primary_prevalence[[task]])) {
            plot = plot + geom_vline(xintercept = primary_prevalence[[task]],
                                     linetype = "dashed",
                                     color = "#6baed6"
                                     ) + 
                          annotate("text",
                                   x = primary_prevalence[[task]],
                                   y = min(Y_MAX, max(df[[metric]])),
                                   label = paste0("Primary care (", 100*round(primary_prevalence[[task]], 3), "%)"),
                                   color = "#6baed6",
                                   angle = 90,
                                   vjust = -0.5, hjust = 1.0
                                   )
            interpolation_target = primary_prevalence[[task]]
        }
        
        # Mark interpolated values for specific panels
        x = lapply(unique(df$Panel), function(target) interpolate_values(df, panel=target, interpolation_target, metric=metric))
        interpolated = data.frame(Panel = unique(df$Panel), NNP = as.numeric(x))
        # Add a X% plotsize gap between primary-care-vline and the text
        interpolated$x = interpolation_target + my_maxp/12

        df_labels <- interpolated %>%
          arrange(NNP) %>%
          mutate(y = 1.5*(1:n()) + max(NNP) - 0.8, nnp_label = paste0(" ", str_replace(Panel, "==>", "\u2192"), " = ", round(NNP, 2)))
        # Add singe or two stage
        df_labels$PanelType = ifelse(grepl("==> TE", df_labels$Panel), "With TE followup", "Standalone test")
        
        print(as.data.frame(df_labels))
        
        # Add the plot elements on top

        plot = plot + 
                    geom_line() +
                    scale_color_manual(values = myColors) +
                    scale_fill_manual(values = myColors) +
                    theme_pubr() +
                    ylim(0, y_max) +
                    geom_segment(data = df_labels, aes_string(x = "x", y = "y", xend = primary_prevalence[[task]], yend = "NNP"), linetype = "dotted") +
                    geom_label(data = df_labels, 
                               aes(x = x, y = y, label = nnp_label), 
                               fill="white",
                               label.size = NA, # disable this for two-stage NNP plots
                               hjust = 0)

        
        plots[[task]] = plot + theme(legend.position = "none")

    }
    
    return(plots)
    
}

```


# Read in performance and prevalence rates and choose the biomarker panels to plot

```{r prepare-data}
library(dplyr)
library(stringr)
library(purrr)

panel_replace = c("Plasma extra-cellular matrix" = "Plasma ECM proteins",
                  "Host-microbial multi-omics" = "Full multi-omics",
                  "Te-reference" = "Gold standard",
                  "Blep-reference" = "Gold standard",
                  "Ast-reference" = "Gold standard",
                  "Gold-standard (blep)" = "Gold standard",
                  "Gold-standard (ast)" = "Gold standard",
                  "Gold-standard (te >=8)" = "Gold standard",
                  "Gold-standard (te >=12)" = "Gold standard",
                  "Gold-standard" = "Gold standard",
                  "Plasma proteomics" = "Plasma proteins",
                  "Plasma lipidomics" = "Plasma lipids",
                  "Plasma metabolomics" = "Plasma metabolites")
pheno_replace = c("S>=2" = "Steatosis",
                  "S>=1" = "Steatosis",
                  "I>=2" = "Inflammation",
                  "F>=2" = "Significant fibrosis",
                  "F>=3" = "Advanced fibrosis")

prevalences   = read.csv("prevalences.txt", header=TRUE, sep="\t", fill=TRUE) %>%
                  mutate(Prevalence = Disease/Total)

performances = read.csv("performances.txt", header=TRUE, sep="\t", fill=TRUE)
performances = performances %>%
                  mutate(phenotype = str_replace_all(phenotype, fixed(pheno_replace))) %>%
                  mutate(omic.name = str_to_sentence(omic.name)) %>%
                  mutate(Panel := str_replace_all(omic.name, fixed(panel_replace))) %>%
                  #filter(! Panel %in% c("Host multi-omics")) %>%
                  mutate(NNP = 1/(PPV+NPV-1))
  
# Fake 'set' column for newer tables from Camila
performances$set = "test"

# Step 1: Filter the data.frame
top_nnp_df = performances %>%
                filter(set == "test" & TNR > 0) %>%
                filter(NNP > 0) %>%
                arrange(NNP) %>%
                group_by(phenotype) %>%
                slice_head(n = 4) %>%
                ungroup()

# Step 2: Filter the data.frame for specific panels
specific_panels_df_1 = performances %>%
                        filter(Panel %in% c("Plasma proteins", "Clinical features") & set == "test" & TNR > 0)
specific_panels_df_2 = performances %>%
                        filter(Panel %in% c("Gold standard") & set == "test" & TNR > 0)
specific_panels_df_3 = performances %>%
                        filter(Panel %in% c("Plasma metabolomics") & phenotype == 'Advanced fibrosis' & set == "test" & TNR > 0)

# Combine the two data.frames and remove duplicates
combined_df = bind_rows(top_nnp_df, specific_panels_df_1, specific_panels_df_2, specific_panels_df_3) %>%
                distinct()

# Step 3: Select unique pairs of 'phenotype' and 'panel'
unique_pairs = combined_df %>%
                  select(phenotype, Panel) %>%
                  distinct() %>%
                  mutate(Panel = ifelse(phenotype == "Advanced fibrosis" & Panel == "Plasma targeted-metabolomics", "Plasma lipids", Panel))

# Step 4: Subset the entire data.frame to keep only these pairs
# and remove selected panels
performances = performances %>%
                  semi_join(unique_pairs, by = c("phenotype", "Panel")) %>%
                  filter(!(phenotype == "Significant fibrosis" & Panel == "Full multi-omics")) %>% # Getting crowded
                  filter(!(phenotype == "Inflammation" & Panel == "Host multi-omics")) %>% # Getting crowded
                  filter(!(phenotype == "Advanced fibrosis" & Panel == "Host multi-omics")) # same as plasma proteomics

```

## Plot the biomarker panels

Also assign target primary and secondary care prevalences to highlight in the plot.
```{r diagnosis-test, fig.height=5.66, fig.width=8}
# Load ggplot2
library(ggplot2)
library(ggpubr)
library(scales)
library(RColorBrewer)
library(cowplot)

primary_prevalence = list("Inflammation" = 0.20,
                       "Significant fibrosis" = 0.15,
                       "Advanced fibrosis" = 0.03,
                       "Steatosis" = 0.30
                      )
secondary_prevalence = list("Inflammation" = NA,
                       "Significant fibrosis" = 0.40,
                       "Advanced fibrosis" = 0.20,
                       "Steatosis" = NA
                      )


# Get the default ggplot2 color palette
panels = c("Plasma cytokines", "Clinical features", "Plasma proteins", "Host multi-omics", "Plasma lipids", "Gold standard", "Plasma metabolites")
myColors = brewer.pal(length(panels), "Dark2")  # Choose a palette and number of colors
names(myColors) = panels  # Assign names to the colors

# Make the actual plots
metric = "NNP"
plots = plot_metric_trends(performances, prevalences=prevalences, primary_prevalence=primary_prevalence, secondary_prevalence=secondary_prevalence, Y_MAX=20, MAX_P=0.25, metric=metric)
merged_plot = plot_grid(plots[["Steatosis"]], plots[["Inflammation"]], plots[["Significant fibrosis"]], plots[["Advanced fibrosis"]], ncol = 2, nrow = 2)
merged_plot
ggsave(paste("diagnostic_markers", metric, "pdf", sep="."), plot=merged_plot, device=cairo_pdf, height=1.7*8/1.618, width=1.7*8, units="in", dpi=300)

#plots[["Significant fibrosis"]]
```

## Plot two-stage diagnosis for Significant fibrosis

```{r diagnosis-two-stage, fig.height=8/1.618, fig.width=8}
# Load ggplot2
library(ggplot2)
library(scales)
library(dplyr)
library(RColorBrewer)
library(cowplot)
library(stringr)
library(data.table)

#    TPR = TP / (TP + FN)  # True Positive Rate
#    TNR = TN / (FP + TN)  # True Negative Rate

prevalences2   = read.csv("two_stage_prevalences.txt", header=TRUE, sep="\t", fill=TRUE) %>%
                  mutate(Prevalence = Disease/Total)

performances2 = read.csv("two_stage_performances.txt", header=TRUE, sep=",", fill=TRUE)
performances2 = performances2 %>%
                  mutate(phenotype = str_replace_all(phenotype, pheno_replace)) %>%
                  mutate(Panel := str_replace_all(omic.name, panel_replace)) %>%
                  filter(!Panel %like% "Full multi-omics")

# study_prevalence2 = list("Significant fibrosis" = 25/63,
#                        "Advanced fibrosis" = 14/77
#                       )
 primary_prevalence2 = list("Significant fibrosis" = 0.15,
                        "Advanced fibrosis" = 0.03
                       )

# Get the default ggplot2 color palette

panels = c("Plasma cytokines", "x", "Plasma proteins", "Host multi-omics", "x", "FIB-4", "x")
myColors = brewer.pal(length(panels), "Dark2")  # Choose a palette and number of colors
names(myColors) = panels  # Assign names to the colors

# Make the actual plots
plots = plot_metric_trends(performances2, prevalences=prevalences2, primary_prevalence=primary_prevalence2, secondary_prevalence=NULL, Y_MAX=20, MAX_P=0.5)

plots[["Significant fibrosis"]]

ggsave("two_stage_sig_fib.pdf", plot=plots[["Significant fibrosis"]], device=cairo_pdf, height=8/1.618, width=8, units="in", dpi=600)
ggsave("two_stage_sig_fib.png", plot=plots[["Significant fibrosis"]], height=8/1.618, width=8, units="in", dpi=1200)

```

```{r prognosis, fig.asp=21/29.7, fig.width=8}
# Load ggplot2
library(ggplot2)
library(scales)
library(dplyr)
library(RColorBrewer)

params = list()
params[["All-cause mortality"]] =  list(
                                c("Clinical features", 0.7097, 0.7395),
                                #c("Plasma cytokines", 0.7458, 0.6706),
                                c("TE-optimized", 0.7097, 0.7601),
                                #c("Plasma metabolomics", 0.6542, 0.7186),
                                c("Plasma proteomics", 0.7839, 0.6549)
                              )
params[["Decompensation"]] =  list(
                                c("Clinical features", 0.7843, 0.8055),
                                c("Plasma proteomics", 0.872, 0.8143),
                                c("Plasma cytokines", 0.8256, 0.8226),
                                #c("Plasma metabolomics", 0.7265, 0.8257),
                                c("TE-optimized", 0.8431, 0.8711)
                              )
params[["Infection"]] =  list(
                                c("Clinical features", 0.6648, 0.5687),
                                c("Plasma proteomics", 0.6698, 0.6745),
                                c("Plasma cytokines", 0.5714, 0.6878),
                                c("Plasma metabolomics", 0.652, 0.6772),
                                c("TE-optimized", 0.6204, 0.7823)
                              )


study_prevalence = list("All-cause mortality" = 62/333,
                       "Decompensation" = 51/307,
                       "Infection" = 108/255
                      )
primary_prevalence = list("All-cause mortality" = 0.2,
                       "Decompensation" = 0.2,
                       "Infection" = 0.2
                      )

# Get the default ggplot2 color palette
panels = c(panels = c("Host-microbial multi-omics", "Clinical features", "TE-optimized", "Plasma cytokines", "Plasma metabolomics", "Plasma proteomics", "Plasma lipidomics", "FIB-4"))
myColors = brewer.pal(length(panels), "Dark2")  # Choose a palette and number of colors
names(myColors) = panels  # Assign names to the colors

# Make the actual plots
plots = plot_metric_trends(params, study_prevalence=study_prevalence, Y_MAX=10, metric="NNP")
merged_plot = plot_grid(plots[["All-cause mortality"]], plots[["Decompensation"]], plots[["Infection"]], ncol = 2, nrow = 2)
merged_plot

```
