---
title: "Performance of ALD biomarkers as a function of prevalence rate"
output: html_document
date: "2024-10-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Define some functions for biomarker evaluation


## Estimate metrics from truth table

```{r calculate_metrics}

calculate_metrics <- function(TP, TN, FP, FN, metrics = c("FPR", "FNR", "PPV", "NPV", "PSI", "NNP", "NND")) {

    df = data.frame(matrix(ncol = length(metrics) + 1, nrow = 0))
    
    # Calculate PPV and NPV
    PPV = TP / (TP + FP)  # Positive Predictive Value
    NPV = TN / (TN + FN)  # Negative Predictive Value
    
    # Calculate Sensitivity and Specificity
    TPR = TP / (TP + FN)  # True Positive Rate
    TNR = TN / (TN + FP)  # True Negative Rate
    FPR = 1 - TNR
    FNR = 1 - TPR
    
    # Derive NNX
    PSI = PPV + NPV - 1
    NNP = 1/PSI
    J   = TPR + TNR - 1
    NND = 1/J
    
    # Print the metrics
    #metrics = c("P", "TP", "FN", "N", "TN", "FP")
    #cat(paste0("PREV: ", sprintf("%.4f", prevalence), "; ", paste(metrics, sprintf("%.4f", sapply(metrics, function(x) get(x))), sep = ": ", collapse = "; "), "\n"))
    df = rbind(df, as.vector(sapply(metrics, function(x) get(x))))

    names(df) = c(metrics)
    #print(df)
    return(df)
}
```

## Given TPR and TNR, create truth table and call $calculate_metrics$ above for given population size and prevalence

```{r estimate_performance}
estimate_performance <- function(TPR, TNR, prevalence, n_pop = 1e6, metrics = c("FPR", "FNR", "PPV", "NPV", "PSI", "NNP", "NND")) {

    df = data.frame(matrix(ncol = length(metrics), nrow = 0))
    
    P = round(n_pop*prevalence)  # Actual Positives
    N = n_pop - P  # Actual Negatives
    
    # Confusion matrix values
    TP = round(TPR*P)  # True Positives
    TN = round(TNR*N)  # True Negatives
    FN = P - TP  # False Positives
    FP = N - TN  # False Negatives
    
    df = calculate_metrics(TP, TN, FP, FN, metrics=metrics)
    df$PREVALENCE = prevalence
    #print(df)
    return(df)
}
```


## Given TPR and TNR, calculate metrics for a range of prevalences

```{r simulate_estimate}
simulate_estimate <- function(panel, TPR, TNR, n_pop = 1e6, maxp = 0.5, metrics = c("FPR", "FNR", "PPV", "NPV", "PSI", "NNP", "NND")) {

    # Number of estimated NNPs for regression
    N = 60
    
    # Arithmetic series
    
    Astart = 0.10
    Aend = maxp
    
    prev_a = seq(from = Astart, to = Aend, length.out = N)
    
    # Geometric series
    
    Gstart = 1e-4
    Gend = 0.1
    # 1e-3 to 0.5, in 20 steps
    # multiplier = (0.5/1e-3)^(1/20) = 1.364421
    Gmult = (Gend/Gstart)^(1/N)
    prev_g = Gstart*Gmult^(0:(N-1))
    
    # Merge series or pick one
    spectrum = c(prev_g, prev_a)
    #spectrum = prev_g
    
    df = data.frame(matrix(ncol = length(metrics) + 1, nrow = 0))
    
    #spectrum = c(1e-3, 1e-2, 1e-1, 5e-1)
    df = NULL
    for (prevalence in spectrum) {
      if (is.null(df)) {
        df = estimate_performance(TPR, TNR, prevalence, n_pop, metrics=metrics)
      } else {
        df = rbind(df, estimate_performance(TPR, TNR, prevalence, n_pop, metrics=metrics))
      }
    }
    df$Panel = panel
    #print(df)
    return(df)
}

```

## Helper functions for plotting the trends

```{r plotting-helpers}

# Interpolate the NNP value at specific values
interpolate_values <- function(df, panel, prevalence, metric="NNP", MAX_P=0.5) {
  df2 = df %>% filter(Panel == panel)
  #print(df)
  #print(paste("filtering for ", panel))
  #print(df2)
  a = approx(df2[["PREVALENCE"]], df2[[metric]], xout = prevalence)
  ret = c(as.numeric(a$x), as.numeric(a$y))
  #print(ret)
  return(ret)
}

# Function to calculate the 95% confidence interval using percentiles
calculate_ci_nonparametric <- function(data, metric) {
  lower <- quantile(data[[metric]], 0.025)
  upper <- quantile(data[[metric]], 0.975)
  data.frame(
    Panel = data$Panel[1],
    PREVALENCE = data$PREVALENCE[1],
    CI_lower = lower,
    CI_upper = upper
  )
}

# Function to calculate the 95% confidence interval using SD
calculate_ci <- function(data, metric) {
  mean_metric = mean(data[[metric]])
  ci <- 1.96 * sd(data[[metric]]) / sqrt(nrow(data))
  data.frame(
    PREVALENCE = data$PREVALENCE[1],
    Panel = data$Panel[1],
    CI_lower = mean_metric - ci,
    CI_upper = mean_metric + ci,
    CI_mean  = mean_metric
  )
}
```

## Actual function plotting the trends

```{r plotting}
# Make a list of ggplot objects
plot_metric_trends <- function(performances, plot_targets, prevalences=NULL, primary_prevalence=NULL, secondary_prevalence=NULL, metric="NNP", Y_MAX=20, MAX_P=0.5, draw_cv_ci=0) {

    tasks = unique(performances$phenotype)
    
    # Plot the trends
    plots = list()
    for (task in tasks) {
        my_maxp = MAX_P
        study_prevalence = NA
        if (!is.null(prevalences)) {
          x = prevalences %>%
            filter(phenotype == task & set == 'test')
          study_prevalence = x$Prevalence
          my_maxp = max(MAX_P, ceiling(study_prevalence*10)/10)
        }
        
        ####################
        # Test performances
        ####################
        
        test_params = performances %>%
                        filter(phenotype == task & set == 'test') %>%
                        select(Panel, TPR, TNR)
        test_df = do.call(rbind, pmap(list(test_params$Panel, test_params$TPR, as.numeric(test_params$TNR)), function(x, y, z) simulate_estimate(x, y, z, maxp=my_maxp)))

        y_max = ceiling(max(test_df[[metric]]))
        if (y_max > Y_MAX) {
            y_max = Y_MAX
        }

        # Add singe or two stage
        test_df$PanelType = ifelse(grepl("==> TE", test_df$Panel), "With TE followup", "Standalone test")

        ####################
        # CV performances
        ####################
        # Add CV information
        if (draw_cv_ci == 1) {
            cv_params = performances %>%
                          filter(phenotype == task & set == 'CV') %>%
                          select(Panel, TPR, TNR)
            
            # Use purrr::pmap to iterate over each row and call simulate_estimate
            result_list = pmap(list(cv_params$Panel, cv_params$TPR, as.numeric(cv_params$TNR)), function(x, y, z) simulate_estimate(x, y, z, maxp=my_maxp))
            
            # Combine the results into a single data.frame
            cv_df = bind_rows(result_list)
            
            
            # Calculate the confidence intervals
            ci_df = cv_df %>%
              group_by(Panel, PREVALENCE) %>%
              do(calculate_ci(., metric)) %>%
              ungroup()
    
            # Merge test performance and CV confidence intervals
            df = test_df %>%
                    inner_join(ci_df, by = c("Panel", "PREVALENCE"))
            
            #######################
            # What are we plotting?
            #######################

            df = df %>%
                  mutate(!!sym(metric) := CI_mean)
        } else {
          df = test_df
          df$CI_lower = 0
          df$CI_upper = 5
        }
        
        # Create the plot

        # Legend for linetype only if there are multiple test types
        if (length(unique(df$PanelType)) > 1) {
            plot = ggplot(df, aes(x = PREVALENCE,
                              y = get(metric),
                              ymin = CI_lower,
                              ymax = CI_upper,
                              linetype = PanelType)
                          ) +
                        labs(title = task,
                             x = "Prevalence",
                             y = metric,
                             color = "Biomarker panel",
                             linetype = "Test type"
                          ) +
                        scale_linetype_manual(values = c("With TE followup" = "solid",
                                                         "Standalone test" = "dashed"))
        } else {
            plot = ggplot(df, aes(x = PREVALENCE,
                              y = get(metric),
                              ymin = CI_lower,
                              ymax = CI_upper)
                          ) +
                        labs(title = task,
                             x = "Prevalence",
                             y = metric,
                             color = "Biomarker panel"
                             )
        }
        
        # Add other elements

        plot = plot + 
                    geom_line(aes(color = gsub(" ==> TE", "", Panel))) +
                    scale_color_manual(values = myColors) +
                    scale_fill_manual(values = myColors) +
                    theme_pubr() +
                    theme(
                      legend.position = c(0.74, 0.95),  # Position in the top right corner
                      # 0.74,0.95 for NNP-diagnostics
                      legend.justification = c("right", "top"),  # Align the legend box
                      legend.background = element_blank(),  # Box around the legend
                      legend.box.background = element_rect(color = "black"),  # Box around the legend box
                      legend.text = element_text(size = 8)  # Reduce the font size
                    ) +
                    ylim(0, y_max)
        
        if (draw_cv_ci == 1) {
          plot = plot + geom_ribbon(aes(fill = gsub(" ==> TE", "", Panel)), color = NA, alpha = 0.2, show.legend = FALSE)

        }
        
        # Add vertical lines for prevalences
        # Also mark what should be interpolation target
        interpolation_target = NULL
        if (!is.null(prevalences)) {
            plot = plot + geom_vline(xintercept = study_prevalence,
                                    linetype = "dotted",
                                    color = "#08519c"
                                    ) +
                          annotate("text",
                                   x = study_prevalence,
                                   y = min(Y_MAX, max(df[[metric]])),
                                   label = paste0("This study (", 100*round(study_prevalence, 3), "%)"),
                                   color = "#08519c",
                                   angle = 90,
                                   vjust = -0.5, hjust = 1.0
                           )
            interpolation_target = study_prevalence
        }
        # secondary is plotted before primary, because we want primary to have preference for showing interpolated values
        if (!is.null(secondary_prevalence) && !is.na(secondary_prevalence[[task]])) {
            plot = plot + geom_vline(xintercept = secondary_prevalence[[task]],
                                     linetype = "dotted",
                                     color = "#3182bd"
                                     ) + 
                          annotate("text",
                                   x = secondary_prevalence[[task]],
                                   y = min(Y_MAX, max(df[[metric]])),
                                   label = paste0("Secondary care (", 100*round(secondary_prevalence[[task]], 3), "%)"),
                                   color = "#3182bd",
                                   angle = 90,
                                   vjust = -0.5, hjust = 1.0
                                   )
            interpolation_target = primary_prevalence[[task]]
        }
        if (!is.null(primary_prevalence) && !is.na(primary_prevalence[[task]])) {
            plot = plot + geom_vline(xintercept = primary_prevalence[[task]],
                                     linetype = "dotted",
                                     color = "#6baed6"
                                     ) + 
                          annotate("text",
                                   x = primary_prevalence[[task]],
                                   y = min(Y_MAX, max(df[[metric]])),
                                   label = paste0("Primary care (", 100*round(primary_prevalence[[task]], 3), "%)"),
                                   color = "#6baed6",
                                   angle = 90,
                                   vjust = -0.5, hjust = 1.0
                                   )
            interpolation_target = primary_prevalence[[task]]
        }
        
        # Mark interpolated values for specific panels
        x = sapply(plot_targets[[task]], function(target) interpolate_values(df, panel=target, interpolation_target, metric=metric))
        interpolated = data.frame(x)
        row.names(interpolated) = c("x", "y")
        names(interpolated) = plot_targets[[task]]
        # Reorder based on 'y'
        interpolated = interpolated[, order(as.numeric(interpolated["y", ]), decreasing = TRUE)]
        
        vjustx = c(-0.2, -0.1, 1.1, 1.2, rep(1.2, 10))
        hjustx = c(-0.2, 1.2, 1.2, -0.2, rep(-0.2, 10))
        for (i in 1:length(names(interpolated))) {
            target = names(interpolated)[i]
            x1 = interpolated[[target]][1]
            y1 = interpolated[[target]][2]
            plot = plot + annotate("point", x = x1, y = y1, color = myColors[[gsub(" ==> TE", "", target)]])
            plot = plot + annotate("text", x = x1, y = y1, label = round(y1, 2), color = myColors[[gsub(" ==> TE", "", target)]], vjust = vjustx[i], hjust = hjustx[i])
        }
        
        plots[[task]] = plot
    }
    
    return(plots)
    
}

```

# Read in performance and prevalence rates and choose the biomarker panels to plot

```{r prepare-data}
library(dplyr)
library(stringr)

panel_replace = c("Plasma extra-cellular matrix" = "Plasma ECM proteins",
                  "Host-microbial multi-omics" = "Full multi-omics",
                  "Te-reference" = " Gold standard",
                  "Blep-reference" = " Gold standard",
                  "Ast-reference" = " Gold standard",
                  "Gold-standard (blep)" = " Gold standard",
                  "Gold-standard (ast)" = " Gold standard",
                  "Gold-standard (te >=8)" = " Gold standard",
                  "Gold-standard (te >=12)" = " Gold standard",
                  "Plasma proteomics" = "Plasma proteins",
                  "Plasma lipidomics" = "Plasma lipids",
                  "Plasma metabolomics" = "Plasma metabolites")
pheno_replace = c("S>=2" = "Steatosis",
                  "S>=1" = "Steatosis",
                  "I>=2" = "Inflammation",
                  "F>=2" = "Mild fibrosis",
                  "F>=3" = "Advanced fibrosis")

prevalences   = read.csv("prevalences.txt", header=TRUE, sep="\t", fill=TRUE) %>%
                  mutate(Prevalence = Disease/Total)

performances = read.csv("performances.txt", header=TRUE, sep="\t", fill=TRUE)
performances = performances %>%
                  mutate(phenotype = str_replace_all(phenotype, fixed(pheno_replace))) %>%
                  mutate(omic.name = str_to_sentence(omic.name)) %>%
                  mutate(Panel := str_replace_all(omic.name, fixed(panel_replace))) %>%
                  filter(! Panel %in% c("Host multi-omics"))
  
# Fake 'set' column for newer tables from Camila
performances$set = "test"

# Step 1: Filter the data.frame
top_nnp_df = performances %>%
                filter(set == "test" & TNR > 0) %>%
                filter(NNP > 0) %>%
                arrange(NNP) %>%
                group_by(phenotype) %>%
                slice_head(n = 4) %>%
                ungroup()

# Step 2: Filter the data.frame for specific panels
specific_panels_df_1 = performances %>%
                        filter(Panel %in% c("Plasma proteins", "Clinical features") & set == "test" & TNR > 0)
specific_panels_df_2 = performances %>%
                        filter(Panel %in% c(" Gold standard") & set == "test" & TNR > 0)
specific_panels_df_3 = performances %>%
                        filter(Panel %in% c("Plasma metabolomics") & phenotype == 'Advanced fibrosis' & set == "test" & TNR > 0)

# Combine the two data.frames and remove duplicates
combined_df = bind_rows(top_nnp_df, specific_panels_df_1, specific_panels_df_2, specific_panels_df_3) %>%
                distinct()

# Step 3: Select unique pairs of 'phenotype' and 'panel'
unique_pairs = combined_df %>%
                  select(phenotype, Panel) %>%
                  distinct() %>%
                  mutate(Panel = ifelse(phenotype == "Advanced fibrosis" & Panel == "Plasma targeted-metabolomics", "Plasma lipids", Panel))

# Step 4: Subset the entire data.frame to keep only these pairs
performances = performances %>%
                  semi_join(unique_pairs, by = c("phenotype", "Panel"))

```

## Plot the biomarker panels

Also ssign target primary and secondary care prevalences to highlight in the plot.

```{r diagnosis-test, fig.height=5.66, fig.width=8}
# Load ggplot2
library(ggplot2)
library(ggpubr)
library(scales)
library(RColorBrewer)
library(cowplot)

primary_prevalence = list("Inflammation" = 0.20,
                       "Mild fibrosis" = 0.15,
                       "Advanced fibrosis" = 0.03,
                       "Steatosis" = 0.30
                      )
secondary_prevalence = list("Inflammation" = NA,
                       "Mild fibrosis" = 0.40,
                       "Advanced fibrosis" = 0.20,
                       "Steatosis" = NA
                      )


# Get the default ggplot2 color palette
panels = c("Plasma cytokines", "Clinical features", " Gold standard", "Full multi-omics", "Plasma lipids", "Plasma proteins", "Plasma metabolites")
myColors = brewer.pal(length(panels), "Dark2")  # Choose a palette and number of colors
names(myColors) = panels  # Assign names to the colors

# values to be highlighted
plot_targets = list(
                    "Mild fibrosis" = c("Full multi-omics", "Plasma cytokines", " Gold standard"),
                    "Advanced fibrosis" = c("Clinical features", "Plasma proteins", " Gold standard", "Plasma lipids"),
                    "Inflammation" = c("Full multi-omics", "Plasma cytokines", " Gold standard"),
                    "Steatosis" = c("Plasma cytokines", "Plasma proteins", "Plasma metabolites", " Gold standard")
                    )


# Make the actual plots
metric = "NNP"
plots = plot_metric_trends(performances, plot_targets=plot_targets, prevalences=prevalences, primary_prevalence=primary_prevalence, secondary_prevalence=secondary_prevalence, Y_MAX=20, MAX_P=0.5, metric=metric)
merged_plot = plot_grid(plots[["Steatosis"]], plots[["Inflammation"]], plots[["Mild fibrosis"]], plots[["Advanced fibrosis"]], ncol = 2, nrow = 2)
ggsave(paste("Figure4", metric, "pdf", sep="."), plot=merged_plot, height=1.7*8/1.618, width=1.7*8, units="in", dpi=300)

```


## Plot two-stage diagnosis for mild fibrosis

```{r diagnosis-two-stage, fig.asp=21/29.7, fig.width=6}
# Load ggplot2
library(ggplot2)
library(scales)
library(dplyr)
library(RColorBrewer)
library(cowplot)

prevalences2   = read.csv("two_stage_prevalences.txt", header=TRUE, sep="\t", fill=TRUE) %>%
                  mutate(Prevalence = Disease/Total)

performances2 = read.csv("two_stage_performances.txt", header=TRUE, sep=",", fill=TRUE)
performances2 = performances2 %>%
                  mutate(phenotype = str_replace_all(phenotype, pheno_replace)) %>%
                  mutate(Panel := str_replace_all(omic.name, panel_replace)) %>%
                  filter(! Panel %in% c("Host multi-omics"))

primary_prevalence2 = list("Mild fibrosis" = 0.15,
                        "Advanced fibrosis" = 0.03
                       )

panels = c("Plasma cytokines", "x", "x", "Full multi-omics", "x", "Plasma proteins", "x", "FIB-4")
myColors = brewer.pal(length(panels), "Dark2")  # Choose a palette and number of colors
names(myColors) = panels  # Assign names to the colors

plot_targets = list(
                    #"Advanced fibrosis" = c("Clinical features", "Plasma proteomics", "TE-optimized"),
                    "Mild fibrosis" = c("Full multi-omics ==> TE", "Plasma cytokines ==> TE", "FIB-4 ==> TE", "FIB-4")
                    )

# Make the actual plots
plots = plot_metric_trends(performances2, plot_targets=plot_targets, prevalences=prevalences2, primary_prevalence=primary_prevalence2, secondary_prevalence=NULL, Y_MAX=20, MAX_P=0.5)

ggsave("FigureS14.pdf", plot=plots[["Mild fibrosis"]], height=8/1.618, width=8, units="in", dpi=300)

```

